---
title: "Mapping source year for spotted owl predicted habitat"
author: "Last updated by Jordan Wingenroth on "
date: "`r format(Sys.time(), '%D')`"
output: github_document
---

## Environment info

Code excluded here for brevity.

```{r echo = FALSE}

knitr::opts_chunk$set(fig.width=12, fig.height=8, fig.align = "center")

# Load R packages
library(tidyverse)
library(curl)
library(sf)
library(spData) # Background maps

sf_use_s2(FALSE) # Turn off spherical geometry

```

## When were the data informing spotted owl habitat preference collected?

Our goal was to make sure that the dataset we use to predict spotted owl habitat preference is not informed by landscape changes caused by the Rim Fire.

Data were sourced [here](https://map.dfg.ca.gov/metadata/ds2185.html).

Some legwork was required because of the formatting of the data. I was unable to access the geospatial data in R initially, and when I tried GIS, I was unable to use the other variables. I converted the raster to a vector file in GIS, dropping the other attributes, and now I'll join them back up in R. I clipped to the area around the Rim Fire for the conversion because converting the whole state of California took too long. Ultimately, all that was needed in R was a test for the latest year in this subset.

```{r}

owl_vat <- st_read("./data/owl_raster/ds2185.gdb", layer = "VAT_ds2185") %>%
    tibble()

owl_shp <- st_read("./data/owl_raster/owl_raster_key.shp")

owl_vat %>%
    filter(Value %in% owl_shp$VALUE) %>%
    mutate(year = as.numeric(SOURCE_YEAR)) %>%
    summarise(max(year))

```

A value of less than 2013 indicates that we don't need to worry about the Rim Fire affecting our spotted owl habitat preference data.

I'll verify that the subset of the owl data covers the vicinity of the Rim Fire. Doing anything with the super-high-resolution `owl_shp` point data takes forever so I am going to subsample it randomly to get a general idea of its spatial range.

```{r}

fires <- st_read("./data/CA_fire_perimeters/California_Fire_Perimeters__all_.shp")

rim <- fires %>%
    filter(FIRE_NAME == "RIM", YEAR_ == 2013)

owl_shp[sample(1:nrow(owl_shp), 1e4),] %>%
    ggplot() +
    geom_sf() +
    geom_sf(data = rim, fill = "red", alpha = .5) +
    theme_bw()

```
